Computational methods in landscape ecology

### 2024-09-20 Sections/01-introduction.qmd
Comments JN

Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include include but are not limited to pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

>>>>

Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include include but are not limited to pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, including human activities [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of landscape ecology experiments].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

### 2024-09-20 Sections/02-spatial-patterns.qmd
Comments JN

Traditionally and at present, most of the prominent approaches that quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
The strength of landscape metrics is that they are uncomplicated to apply and communicate [@Lausch2015UnderstandingQuantifying], however, limitations have also been identified. 
These limitations include shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

To address some limitations, for example correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees have been used to identify core metrics that capture main components of landscape patterns.
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
These components of landscape patterns are in line with revent reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] that have similarly suggested that there are also two fundamental components of landscape patterns, namely amount and adjacency, and that these components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.

Software to calculate landscape metrics include the *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpensource] and *multilandr* [@Huais2024MultilandrPackage] R packages, the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
<!-- MB: Are there any libraries to perform factor analysis in R that are kind of specific to landscape ecology? I have had a draft on how to perform factor analysis with pylandstats https://github.com/martibosch/pylandstats-notebooks/blob/main/notebooks/05-cluster-analysis.ipynb, hopefully I will include it in the next release before -->
<!-- MH: Would not mention PCA or similar here. Thats a general statistical issue and outside scope imho -->

## Landscape Mosaic

The Landscape Mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes.
The approach uses a moving window to determine the proportions of these three classes within each central grid cell.
These proportions are then classified into 19 mosaic classes based on thresholds that describe the presence, dominance, or uniqueness of each class.
This allows the assessment of content, context, and interface zones of LULC data in a scale-dependent manner.
Recently, the method has been improved by expanding the classification to 103 classes and incorporating heatmap visualizations and summaries [@Vogt2024RevisitingLandscape].

@@

R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.

>>>>

Traditionally and at present, most of the prominent approaches that quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
The strength of such landscape metrics is that they are easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, limitations have also been identified. 
These limitations include shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

To address some limitations, for example correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees have been used to identify core metrics that capture main components of landscape patterns.
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
These components of landscape patterns are in line with recent reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] that have similarly suggested that there are also two fundamental components of landscape patterns, namely amount and adjacency.
These components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory, and later formalized in @Nowosad2019InformationTheory.

Software to calculate landscape metrics include the *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpensource] and *multilandr* [@Huais2024MultilandrPackage] R packages, the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].

The landscape mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes [@Riitters2009IndicatorForest].
The approach uses a moving window to determine the proportions of these three classes within each central grid cell.
These proportions are then classified into 19 mosaic classes based on thresholds that describe the presence, dominance, or uniqueness of each class.
This allows the assessment of content, context, and interface zones of LULC data in a scale-dependent manner.
Recently, the method has been improved by expanding the classification to 103 classes and incorporating heatmap visualizations and summaries [@Vogt2024RevisitingLandscape].

@@

R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], or *terra* [@Hijmans2024TerraSpatial] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance], or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for the use in landscape ecology.

### 2024-09-20 Sections/06-genetics.qmd
Comments JN

Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], *landgenreport* [@Gruber2015LandgenreportNew], *adegenet* [@Jombart2008AdegenetPackage], or *ape* [@Paradis2019Ape50], among others.
Furthermore, a series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].

>>>>

Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], *landgenreport* [@Gruber2015LandgenreportNew], *adegenet* [@Jombart2008AdegenetPackage], or *ape* [@Paradis2019Ape50], among others.
Furthermore, a series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., @Brown2017SDMtoolbox20].

### 2024-09-20 Sections/08-discussion.qmd
Comments JN

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
In combination with other increasingly available data sources, e.g, citizen science data, ecology is entering the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance, artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].

Climate change as well as the biodiversity crisis are two intervening and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
However, up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
Nevertheless, the review provided here that showcases current computational methods in landscape ecology will hopefully only be a snapshot in time because of the development and emergence of future analytical approaches.

>>>>

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance, artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].

Climate change as well as the biodiversity crisis are two intervening and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
Up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
Nevertheless, the review provided here that showcases current computational methods in landscape ecology will only be a snapshot in time because of the development and emergence of future analytical approaches.
